---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r,echo=TRUE}
#Downloading the File:
dataurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if (!file.exists("activity.zip")) {
	download.file(dataurl, destfile = "activity.zip", method = "curl")
}

# Unzip data:
datafile <- "activity.csv"
if (!file.exists(datafile)) {
	unzip(zipfile = "activity.zip")
}

# Load data:
data <- read.csv(datafile, header = TRUE)
# To prevent scientific notation:
options(scipen = 100)

#The data will look like this:
dim(data)
str(data)
head(data,5)
summary(data)
```
```{r,warning=FALSE,message=FALSE}
#For processing the data, here I will use dplyr package:
library(dplyr)
#If it is not install use install.packages('dplyr')
```


## What is mean total number of steps taken per day?
```{r,message=FALSE}
#processing data for this question
steps_per_day <- select(data,steps,date,interval) %>% group_by(date) %>% summarise(total_steps=sum(steps,na.rm = TRUE))
mean_total_steps <- mean(steps_per_day$total_steps)
median_total_steps <- median(steps_per_day$total_steps)

#plot
hist(steps_per_day$total_steps, 
     main = "Histogram of total steps per day", 
     xlab = "Total steps per day", 
     ylab = "Frequency [number of days]", 
     breaks = 25,
     col = "lightgreen"
)

#Here, black line shows mean of total number of steps taken per day
abline(v = mean_total_steps, lwd = 1, lty = 2)
#Here, red line shows median of total number of steps taken per day
abline(v = median_total_steps, lwd = 1, lty = 2, col = "red")  

```

#### The mean number of steps taken each day is `r round(mean_total_steps, 0)`, the median number is `r round(median_total_steps, 0)`.  


## What is the average daily activity pattern?
```{r,message=FALSE}
#processing data for this question
steps_by_interval <- select(data, steps, date, interval) %>% 
	group_by(interval) %>% 
	summarize(average_steps = mean(steps, na.rm = TRUE))

plot(steps_by_interval$average_steps,
     type='l',
     main = "Average Step by Interval", 
     ylab = "Average Step", 
     xlab = "Interval", 
     col = "blue"
)
max_average_steps <- max(steps_by_interval$average_steps)
max_average_steps_interval <- steps_by_interval[
	steps_by_interval$average_steps == max_average_steps,
]$interval
abline(v = max_average_steps_interval, lwd = 1, lty = 2, col = "red")
abline(h = max_average_steps, lwd = 1, lty = 2, col = "red")  
```
  
`r max_average_steps_interval` is the 5-minute interval that, on average, contains the maximum number of `r round(max_average_steps)` steps. 

  
## Imputing missing values
The total number of missing values in the dataset is `r sum(is.na(data))`
```{r,message=FALSE}
replaced_data <- data %>% mutate(replace_steps = ifelse(is.na(steps),steps_by_interval[2],steps))

head(replaced_data,1)

# Could not figure out how to take 1 value for a particular row attaching whole vector

#Thought of replacing na values with mean
steps_mm <- select(data,steps,date,interval) %>% group_by(date) %>% summarise(mean_steps=mean(steps,na.rm = TRUE))
steps_mm$mean_steps[is.nan(steps_mm$mean_steps)] <- 0

updated_data <- data %>% mutate(update_steps = ifelse(is.na(steps),steps_mm[2],steps))

head(updated_data,1)
#Getting same as above output

#trying this
updated_activity <- data
for(r in 1:nrow(updated_activity)) {
	if (is.na(updated_activity$steps[r])) {
		i <- updated_activity$interval[r];
		avg <- steps_by_interval[steps_by_interval$interval==i,]$average_steps
		
		updated_activity$steps[r] <- avg
	}
}

head(updated_activity[which(
	updated_activity$date %in% c("2012-10-01", "2012-10-08", "2012-11-01", "2012-11-04", "2012-11-09", "2012-11-10") & 
	updated_activity$interval %in% c(600,1000,2000)),], n = 30)

updated_steps_by_date <- select(updated_activity, steps, date, interval) %>% 
    group_by(date) %>% 
    summarize(total_steps = sum(steps, na.rm = TRUE))

hist(updated_steps_by_date$total_steps, 
	main = "Histogram of total steps per day,\nreplacing NAs with average steps for given interval", 
	xlab = "Total steps per day", 
	ylab = "Frequency [number of days]", 
	breaks = 20,
	border = "cadetblue4",
	col = "cadetblue3"
)
updated_mean_total_steps <- mean(updated_steps_by_date$total_steps)
updated_median_total_steps <- median(updated_steps_by_date$total_steps)
abline(v = updated_mean_total_steps, lwd = 1, lty = 2, col = "red")
abline(v = updated_median_total_steps, lwd = 1, lty = 2, col = "red")

```
  
The mean number of steps taken each day is now `r round(updated_mean_total_steps, 0)` 
(compared to `r round(mean_total_steps, 0)` with NAs), and the median number 
is `r round(updated_median_total_steps, 0)` now (compared to `r round(median_total_steps, 0)` with NAs). 


## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE, fig.width=10}
activity_with_weekday <- mutate(updated_activity, 
	weekday = weekdays(as.Date(updated_activity$date)),
	weekday_is = as.factor(ifelse(weekday=="Saturday" | weekday=="Sunday", "weekend", "weekday"))
)
# Sample of the extended data
head(select(activity_with_weekday, date, weekday, weekday_is) %>% distinct(date), n = 20)
library(lattice) 
# Get average number of steps per weekday/weekend and interval
summary <- select(activity_with_weekday, steps, date, interval, weekday_is) %>% 
    group_by(weekday_is, interval) %>% 
    summarize(average_steps = mean(steps, na.rm = TRUE))
xyplot(average_steps ~ interval | weekday_is, data = summary, layout = c(1,2), type = "l",
	main = "Average steps by 5-minute interval",
	xlab = "Interval",
	ylab = "Average number of steps",
	col = "cadetblue4"
)
```

Yes, there is a difference in activity patterns between weekdays and weekends.
It looks like the activities start a little later and more gradually on the weekends. The step volume is also a bit higher throughout the day.
On weekdays the step volume appears to be higher and increase quicker (more steeply) in the mornings, but remains a little lower throughout the day.

```{r, echo=TRUE}
# Clean up
rm(list=ls(all=TRUE)) 
```
